---
title: "Geom examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geom examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE   # figures require a browser; set TRUE for local preview
)
```

This vignette shows how to use each geometry in highdir.
Every example follows the same two-step workflow:

1. **`hd_spec()`** — describe *what* the data means (columns, labels)
2. **`hd_make()`** — decide *how* to render it (geom, backend, options)

The same `hd_spec` object can be passed to `hd_make()` multiple times with
different geom types, backends, or `hd_opts()` objects — there is no need
to rebuild the spec.

```{r shared-data}
library(highdir)

# ── Time-series dataset (single group) ────────────────────────────────────────
# Annual alcohol consumption estimate with 95 % confidence interval.
alcohol <- data.frame(
  year     = 2010:2023,
  adj_mean = c(6.2, 6.5, 6.8, 7.0, 7.3, 7.1, 6.9, 6.7,
               6.4, 6.2, 5.9, 5.7, 5.5, 5.3),
  lower_95CI = c(5.8, 6.1, 6.3, 6.5, 6.8, 6.6, 6.4, 6.2,
               5.9, 5.7, 5.4, 5.2, 5.0, 4.8),
  upper_95CI = c(6.6, 6.9, 7.3, 7.5, 7.8, 7.6, 7.4, 7.2,
               6.9, 6.7, 6.4, 6.2, 6.0, 5.8)
)

alcohol <- highdir::hd_data1

# ── Time-series dataset (kjonn groups) ──────────────────────────────────────────
alcohol2 <- data.frame(
  year     = rep(2010:2023, each = 2),
  kjonn      = rep(c("Male", "Female"), times = 14),
  adj_mean = c(8.1, 4.5, 8.4, 4.8, 8.7, 5.1, 9.0, 5.3,
               9.2, 5.6, 9.0, 5.4, 8.8, 5.2, 8.6, 5.0,
               8.3, 4.8, 8.1, 4.6, 7.8, 4.3, 7.5, 4.1,
               7.3, 3.9, 7.0, 3.7),
  lower_95CI = c(7.6, 4.0, 7.9, 4.3, 8.2, 4.6, 8.5, 4.8,
               8.7, 5.1, 8.5, 4.9, 8.3, 4.7, 8.1, 4.5,
               7.8, 4.3, 7.6, 4.1, 7.3, 3.8, 7.0, 3.6,
               6.8, 3.4, 6.5, 3.2),
  upper_95CI = c(8.6, 5.0, 8.9, 5.3, 9.2, 5.6, 9.5, 5.8,
               9.7, 6.1, 9.5, 5.9, 9.3, 5.7, 9.1, 5.5,
               8.8, 5.3, 8.6, 5.1, 8.3, 4.8, 8.0, 4.6,
               7.8, 4.4, 7.5, 4.2)
)

alcohol2 <- highdir::hd_data2

# ── Age-group survey dataset (bars and pie) ────────────────────────────────────
survey <- data.frame(
  age_group = rep(c("18-24", "25-34", "35-44", "45-54", "55-64"), each = 2),
  kjonn       = rep(c("Male", "Female"), times = 5),
  pct       = c(42, 38, 55, 61, 48, 52, 60, 57, 65, 70),
  n         = c(120, 115, 200, 210, 180, 175, 160, 155, 140, 145)
)

# ── Category share dataset (pie) ──────────────────────────────────────────────
drinking_freq <- data.frame(
  category = c("Never", "Rarely", "Monthly", "Weekly", "Daily"),
  pct      = c(18, 25, 30, 20, 7)
)
```


## Column chart

Column charts are the default geometry and work with both grouped and
ungrouped data.
The `n` column is shown in the highcharter tooltip alongside the percentage.

```{r column-grouped}
spec_col <- hd_spec(survey,
                    x     = "age_group",
                    y     = "pct",
                    group = "kjonn",
                    n     = "n"
                    )

opts_col <- hd_opts(
  title    = "Alcohol use by age group and kjonn",
  subtitle = "Source: Norwegian Directorate of Health",
  ylim     = c(0, 100),
  yint     = 20,
  ylab  = "Percentage (%)"
)

# Interactive (default)
hd_make(spec_col, "column", opts_col)

# Static ggplot2
hd_make(spec_col, "column", opts_col, backend = "ggplot2")

# Horizontal bars — flip applies to both backends
opts_flip <- hd_opts(
  title    = "Alcohol use by age group and kjonn",
  subtitle = "Source: Norwegian Directorate of Health",
  ylim     = c(0, 100),
  flip     = TRUE
)
hd_make(spec_col, "column", opts_flip)
hd_make(spec_col, "column", opts_flip, backend = "ggplot2")
```


## Line chart

Line charts suit time-series data.
Use `smooth = TRUE` (default) for spline curves, or `FALSE` for straight
segments between points.
`dot_size` controls the marker radius.

```{r line-single}
# Single series — no group column
spec_line1 <- hd_spec(alcohol,
                      x    = "year",
                      y    = "adj_mean"
                      )

opts_line <- hd_opts(
  title    = "Alcohol consumption over time",
  subtitle = "Source: Norwegian Directorate of Health",
  ylim     = c(0, 50),
  ylab = "Litres per capita"
)

# Spline (smooth) — default
hd_make(spec_line1, "line", opts_line)
hd_make(spec_line1, "line", opts_line, backend = "ggplot2")

# Straight segments
hd_make(spec_line1, "line", opts_line, smooth = FALSE)
hd_make(spec_line1, "line", opts_line, smooth = FALSE, backend = "ggplot2")
```

```{r line-grouped}
# Grouped by kjonn
spec_line2 <- hd_spec(alcohol2,
                      x     = "year",
                      y     = "adj_mean",
                      group = "kjonn")

# Larger markers, straight lines
hd_make(spec_line2, "line", opts_line, smooth = FALSE, dot_size = 6)
hd_make(spec_line2, "line", opts_line, smooth = FALSE, dot_size = 6,
        backend = "ggplot2")

# Highcharter only: custom per-group marker shapes
hd_make(spec_line2, "line", opts_line,
        line_symbols = c("circle", "square"))
```


## Arearange chart

Arearange is designed for displaying a central estimate alongside a
confidence interval or min/max band.
`ymin` and `ymax` are **required** — they name the columns that define the
lower and upper bounds of the shaded band.

```{r arearange-single}
# Single series
spec_ar1 <- hd_spec(alcohol,
                    x    = "year",
                    y    = "adj_mean")

opts_ar <- hd_opts(
  title    = "Alcohol consumption with 95% CI",
  subtitle = "Source: Norwegian Directorate of Health",
  ylim     = c(0, 10),
  ylab = "Litres per capita"
)

hd_make(spec_ar1, "arearange", opts_ar,
        ymin = "lower_95CI", ymax = "upper_95CI")

hd_make(spec_ar1, "arearange", opts_ar,
        ymin = "lower_95CI", ymax = "upper_95CI", backend = "ggplot2")
```

```{r arearange-grouped}
# Grouped by kjonn
spec_ar2 <- hd_spec(alcohol2,
                    x     = "year",
                    y     = "adj_mean",
                    group = "kjonn")

hd_make(spec_ar2, "arearange", opts_ar,
        ymin = "lower_95CI", ymax = "upper_95CI")

hd_make(spec_ar2, "arearange", opts_ar,
        ymin = "lower_95CI", ymax = "upper_95CI", backend = "ggplot2")
```


## Pie chart

Pie charts map one categorical column to slice labels and one numeric
column to slice values.
Pass `inner_size = "50%"` (or any CSS percentage) to draw a donut chart.
The `group` column is ignored for pie charts — use `x` for labels and `y`
for values.

```{r pie}
spec_pie <- hd_spec(drinking_freq,
                    x    = "category",
                    y    = "pct")

opts_pie <- hd_opts(
  title    = "Drinking frequency",
  subtitle = "Source: Norwegian Directorate of Health",
  ylab = "Share (%)"
)

# Solid pie
hd_make(spec_pie, "pie", opts_pie)
hd_make(spec_pie, "pie", opts_pie, backend = "ggplot2")

# Donut
hd_make(spec_pie, "pie", opts_pie, inner_size = "50%")
hd_make(spec_pie, "pie", opts_pie, inner_size = "50%", backend = "ggplot2")
```


## Reusing a spec across geoms

Because `hd_spec()` and `hd_opts()` are separate from `hd_make()`, the
same data description can be rendered as any geometry — no code duplication.

```{r reuse}
# One spec, three geoms
spec_shared <- hd_spec(alcohol2,
                       x     = "year",
                       y     = "adj_mean",
                       group = "kjonn")

opts_shared <- hd_opts(
  title    = "Alcohol consumption by kjonn",
  subtitle = "Source: Norwegian Directorate of Health",
  ylim     = c(0, 50),
  ylab  = "Litres per capita"
)

hd_make(spec_shared, "column",    opts_shared)
hd_make(spec_shared, "line",      opts_shared)
hd_make(spec_shared, "arearange", opts_shared,
        ymin = "lower_95CI", ymax = "upper_95CI")
```


## Reusing opts across languages

An `hd_opts()` object only controls presentation.
Create one per language and pair it with the same spec.

```{r bilingual}
opts_no <- hd_opts(
  title    = "Alkoholbruk over tid",
  subtitle = "Kilde: Helsedirektoratet",
  caption  = "Tall om alkohol",
  ylim     = c(0, 50)
)

opts_en <- hd_opts(
  title    = "Alcohol use over time",
  subtitle = "Source: Norwegian Directorate of Health",
  caption  = "Annual health report",
  ylim     = c(0, 50)
)

spec_ts <- hd_spec(alcohol2,
                   x     = "year",
                   y     = "adj_mean",
                   group = "kjonn",
                   ylab  = "Litres per capita")

hd_make(spec_ts, "line", opts_no)
hd_make(spec_ts, "line", opts_en)
```


## Theming

Use `hd_set_theme()` once per session to apply a consistent visual style
across all figures.
Per-figure overrides are set via `hd_opts(hc_theme = ..., gg_theme = ...)`.

```{r theming}
# Session-wide defaults
hd_set_theme(
  hc_theme = "gridlight",
  gg_theme = "classic",
  colors   = c("#025169", "#7C145C", "#C68803")
)

hd_make(spec_ts, "line", opts_en)
hd_make(spec_ts, "line", opts_en, backend = "ggplot2")

# Per-figure override — does not change the session default
hd_make(spec_ts, "line",
        hd_opts(title = "Dark theme", hc_theme = "darkunica"),
        backend = "highcharter")

# Reset to package defaults
hd_set_theme(hc_theme = "default", gg_theme = "classic", colors = NULL)
```


## Saving figures

`hd_save()` exports highcharter figures to HTML or JSON and ggplot2
figures to PNG, SVG, or PDF.
The format is inferred from the file extension.

```{r save, eval = FALSE}
hc_fig <- hd_make(spec_ts, "line", opts_en)
gg_fig <- hd_make(spec_ts, "line", opts_en, backend = "ggplot2")

hd_save(hc_fig, "alcohol_line.html")
hd_save(hc_fig, "alcohol_line.json")

hd_save(gg_fig, "alcohol_line.png")
hd_save(gg_fig, "alcohol_line.svg")
```
