---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{start-use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(highdir)
```


```{r setup, eval = FALSE}

library(highdir)

# Create dataset
df <- data.frame(
  age  = rep(c("18-24", "25-34", "35-44", "45-54"), each = 2),
  sex  = rep(c("Male", "Female"), 4),
  pct  = c(42, 38, 55, 61, 48, 52, 60, 57),
  n    = c(120, 115, 200, 210, 180, 175, 160, 155)
)

# Figure data specification 
spec <- hd_spec(df, x = "age", y = "pct", group = "sex", n = "n")

# Specify figure presentation
opts <- hd_opts(title = "Tall om Report",
                subtitle = "Source: Example data",
                caption  = "Tall om helse",
                ylim = c(0, 80))

hd_make(spec, "column", opts)                       # highcharter (default)
hd_make(spec, "column", opts, backend = "ggplot2")  # static ggplot2
hd_make(spec, "line",   opts, smooth = TRUE)        # smooth spline
hd_make(spec, "pie",    opts)                       # pie / donut

# Disable JavaScript (for static HTML export)
hd_make(spec, "column", opts, use_js = FALSE)

# Save
hd01 <- hd_make(spec, "column")
hd_save(fig = hd01, file = "chart.html")

gg01 <- hd_make(spec, "column", backend = "ggplot2")
hd_save(fig = gg01, file = "chart.png")
```

## Geom 

To list all geom features.

```{r geom}
list_geoms()
```

## Specification Types in highdir

The highdir framework defines four categories of specifications used to
construct plots. Each category serves a distinct role in data definition, visual
configuration, and geom‑level customization.

1. `hd_spec()` 

Defines the data specifications that apply universally across all
geom types. These specifications describe the input data structure and mappings
required by the plotting system.

2. `hd_opts()`

Specifies presentation options that control chart appearance. These options are
shared across multiple geoms and include settings related to layout, style, and
display behavior.

3. Required Specifications

Parameters that a given geom must provide for successful rendering. These vary
by geom and represent the minimum information needed for the geom to function.

4. Optional Specifications

Additional parameters that may be supplied depending on the selected geom. These
extend or refine the behavior of a geom but are not mandatory.

To see the required and optionals arguments:

```{r args, eval = FALSE}
geom_args("line") #args in geom line
geom_args("arearange")

# or
highdir:::.get_geom("ranked_bar")$optional_args
highdir:::.get_geom("arearange")$required_args
```

---

## Theming

```{r, eval = FALSE}
# Set package-wide defaults for the session
hd_set_theme(
  hc_theme = "helsedirektoratet",
  colors   = c("#025169", "#7C145C", "#C68803"),
  font     = "Source Sans Pro"
)

# All subsequent hd_make() calls use these settings automatically
hd_make(spec, "column")
```

---

## JavaScript injection

```{r, eval = FALSE}
fig <- hd_make(spec, "column")

# Inline JS
fig <- hd_add_js(fig, code = "console.log('chart loaded');")

# From a .js file
fig <- hd_add_js(fig, file = "path/to/my-plugin.js")

# From a bundled plugin (inst/js/<name>.js)
fig <- hd_add_js(fig, plugin = "my-plugin")
```
